---
title: "Tutorial on Mantel test in R"
author: "Brian"
date: "5/18/2021"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)

```

#### Purpose

Identify relationship between matrices. For example, it statistically tests whether spacial data are related to  measurement data or not.

#### Required Packages

For conducting Mantel test in R, we need to use a package such as **ade4** or **vegan**. Currently, while **vegan** supports both simple and partial Mantel test, **ade4** supports simple Mantel test only.

Fortunately, regardless of what package you are using, the process is almost identical. The difference is only the name of function.

- vegan : mantel (simple), mantel.partial (partial) 
- ade4 : mantel.rtest (simple)

#### Using **vegan** package:

#### Step 1. Load library

```{r message=FALSE}
library(vegan)
```

#### Step 2. Input data

Load data either from Web or local file.

From Web:

```{r}
ozone <- read.table("https://stats.idre.ucla.edu/stat/r/faq/ozone.csv", sep=",", header=T)
```

From local file. File should be located in the project folder.

```{r}
ozone <- read.table("ozone.csv", sep=",", header=T)
```

These are the top 10 rows of the data we are using for this tutorial.

```{r echo=FALSE}
knitr::kable(
  head(ozone, 10), booktabs = TRUE
)
```

Inspect the structure of data to determine appropriate pre-processing method. It indicates there are 32 rows and 4 columns(Station, Av8top, Lat, Lon). Among columns, Lat and Lon is location data and Av8top is measurement data.

```{r}
str(ozone)
```

Show first 10 rows for quick check.
```{r}
head(ozone, n=10)
```

#### Step 3. Pre-processing

At first, calculate the distance between pairs using **dist** function.

```{r}
# Calculate distance matrix for locations
station.dists <- dist(cbind(ozone$Lon, ozone$Lat)) 

# Calculate distance matrix for measurements
ozone.dists <- dist(ozone$Av8top)

# Control matrix
# Create a matrix arbitrarily for demonstration purpose only
c.dists <-  dist(ozone$Av8top * runif(length(ozone$Av8top)))

```

Then, display data in matrix to verify. (only first 5 rows and columns)

```{r}
# Print matrix
as.matrix(station.dists)[1:5, 1:5]
as.matrix(ozone.dists)[1:5, 1:5]
```


Visualize matrices. 

```{r, echo = FALSE, fig.align = 'center'}
source("Custom_Heatmap.R", local = knitr::knit_global())
p1<-custom_heatmap(as.matrix(station.dists),"Station")
p2<-custom_heatmap(as.matrix(ozone.dists),"Ozone")
grid.arrange(p1, p2, ncol = 2)   

```

#### Step 4. Analysis

Run the Mantel test with 9999 repetitions.

```{r}
# Run simple Mantel test
results <- mantel(station.dists, ozone.dists, permutations = 9999)

# Run partial Mantel test
results_partial <- mantel.partial(station.dists, ozone.dists, c.dists, permutations = 9999)
```

#### Step 5. Ouput

```{r}
# Results of simple Mantel test
show(results)
```

From the results of simple Mantel test, the r is `r results$statistic` and p value is `r results$signif`(less than 0.05). Thus, null hypothesis is rejected and research hypothesis is taken.

```{r}
# Results of partial Mantel test
show(results_partial)
```

From the results of partial Mantel test, the r is `r results_partial$statistic` and p value is `r results_partial$signif`(less than 0.05). Thus, null hypothesis is rejected and research hypothesis is taken.

#### Conclusion

The p value of `r results$pvalue` indicates that the relationship between spacial data and measurement data is statistically significant. Therefore, we can conclude that the location and ozone value are related each other.

#### Using **ade4** package 

#### Step 1. Load library

```{r}
library(ade4)
```

#### Step 2. Input data

Load data either from Web or local file.

From Web:

```{r}
ozone <- read.table("https://stats.idre.ucla.edu/stat/r/faq/ozone.csv", sep=",", header=T)
```

From local file. File should be located in the project folder.

```{r}
ozone <- read.table("ozone.csv", sep=",", header=T)
```

Inspect data structure to determine appropriate pre-processing method. It shows there are 32 rows and 4 columns(Station, Av8top, Lat, Lon). Among columns, Lat and Lon is location data and Av8top is measurement data.

```{r}
str(ozone)
```

Show first 10 rows for quick check.
```{r}
head(ozone, n=10)
```

#### Step 3. Pre-processing

At first, calculate the distance between pair using **dist** function. We prepared three matrices for partial Mantel test.

```{r}
# Calculate distance matrix for locations
station.dists <- dist(ozone[,c('Lon','Lat')]) 

# Calculate distance matrix for measurements
ozone.dists <- dist(ozone$Av8top)
```

Then, display data in matrix to verify. (only first 5 rows and columns)

```{r}
# Print matrix
as.matrix(station.dists)[1:5, 1:5]
as.matrix(ozone.dists)[1:5, 1:5]
```


Visualize matrices. 

```{r, echo = FALSE, fig.align = 'center'}
source("Custom_Heatmap.R", local = knitr::knit_global())
p1<-custom_heatmap(as.matrix(station.dists),"Station")
p2<-custom_heatmap(as.matrix(ozone.dists),"Ozone")
grid.arrange(p1, p2, ncol = 2)   

```

#### Step 4. Analysis

Run the Mantel test with 9999 repetitions.

```{r}
# Run simple Mantel test
results <- mantel.rtest(station.dists, ozone.dists, nrepet = 9999)
```

#### Step 5. Ouput

```{r}
# Results of simple Mantel test
show(results)
```

#### Source code

```{r, eval = FALSE, code=xfun::read_utf8('Tutorial_Mantel_test.R')}
```

#### References

HOW CAN I PERFORM A MANTEL TEST IN R? | R FAQ